---
title: "p8105_hw2_abj2136"
output: github_document
date: "2026-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installing libraries
```{r message = FALSE}
library(tidyverse)
library(readxl)
library(readr)
```


# Problem 2

## Mr Trash Wheel
```{r}
# Looking at the sheets in the file
excel_sheets("202610 Trash Wheel Collection Data.xlsx")

#importing and cleaning the the mr trash wheel dataset

mr_trash_df = read_excel("202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Mr. Trash Wheel",
                      range = "A2:N742",
                      na = c("","NA", ".")
                      ) |>
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(sports_balls = as.integer(round(sports_balls))) |> 
  mutate(year = as.integer(year))
  
  # used glimpse in the console to determine variable types for all three data frames to 
  # determine which variables were entered as characters vs numeric values. In Mr.Trash 
  # the year variable needed to be converted to an integer, and in Professor trash the 
  # plastic bottles variable needed to be converted to numeric. 
```

## Professor Trash Wheel
```{r}
#importing and cleaning the the prof trash wheel dataset

prof_trash_df = read_excel("202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Professor Trash Wheel",
                      range = "A2:M139",
                      na = c("","NA", ".")
                      ) |>
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(plastic_bottles = as.numeric(plastic_bottles))
```

## Gwynnda the Good Wheel of the West
```{r}
#importing and cleaning the the prof trash wheel dataset

gwynnda_df = read_excel("202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Gwynnda the Good Wheel of the W",
                      range = "A2:L394",
                      na = c("","NA", ".")
                      ) |>
  janitor::clean_names() |> 
  drop_na(dumpster)
```

## Merging datasets
```{r}
# creating a trash_wheel variable to keep track of which trash wheel is which after merging

mr_trash_df = mr_trash_df |> 
  mutate(trash_wheel = "Mr. Trash Wheel")

prof_trash_df = prof_trash_df |> 
  mutate(trash_wheel = "Professor Trash Wheel")

gwynnda_df = gwynnda_df |> 
  mutate(trash_wheel = "Gwynnda the Good Wheel of the W")

# merging the datasets

combined_trash_df = 
  bind_rows(
    mr_trash_df,
    prof_trash_df,
    gwynnda_df, 
  )

```

The final combined dataset contains `r nrow(combined_trash_df)` observations and `r ncol(combined_trash_df)` variables, such as `r names(combined_trash_df)`. Professor Trash Wheel collected a total of `r combined_trash_df |> filter(trash_wheel == "Professor Trash Wheel") |>  pull(weight_tons) |> sum(na.rm = TRUE)` tons of trash. Gwynnda the Good Wheel of the West collected `r combined_trash_df |> filter(trash_wheel == "Gwynnda the Good Wheel of the W", str_to_lower(month) == "june", year == 2022) |> pull(cigarette_butts) |> sum(na.rm = TRUE) |> scales::comma()` cigarette butts in June 2022. 


# Problem 3

## Importing and cleaning the baseline demographics data

```{r message = FALSE}

# keeping this here to see what original variables are called before skip

baseline_raw = read_csv("data_mci/MCI_baseline.csv")

# cleaning data set

baseline_df = read_csv("data_mci/MCI_baseline.csv",
                       na = c("", ".", "NA"),
                       skip = 1) |> 
  janitor::clean_names() |> 
  mutate(
    sex = recode_values(sex, 1 ~ "male", 0 ~ "female"),
    apoe4 = recode_values(apoe4, 1 ~ "carrier", 0 ~ "non-carrier")
    ) |> 
    filter(is.na(age_at_onset) | age_at_onset > current_age)
```

The first row of the data set acted like a code book, defining what the values meant for the variables below. I kept one read.csv line with the raw version so I could refer back to this code, as I skipped that row when I was cleaning my data. I had to recode the sex and apoe4 variables to change them from numeric to male/female and carrier/noncarrier. To filter the data to only keep those who had MCI at baseline, I kept everyone who had age of onset = na (never developed MCI) and everyone who developed MCI after baseline (based on if their age of MCI onset was greater than their age at baseline(current_age variable)), as these are the people who would be MCI free at baseline. 

`r nrow(baseline_raw)-1` participants were recruited into the study and `r nrow(baseline_df)` participants were eligible for the study. Of these participants `r baseline_df |> filter(!is.na(age_at_onset)) |> nrow()` developed MCI.The average baseline age among those who participated in the study was `r baseline_df |> pull(current_age) |> mean() |> round (1)` years. `r baseline_df |> filter(sex == "female") |> summarise(prop = mean(apoe4 == "carrier", na.rm = TRUE))|> pull(prop) |> scales::percent(accuracy = 0.1)` of the women in the study are APOE4 carriers.


## Importing and cleaning amyloid data
```{r}

 amyloid_df = read_csv("data_mci/mci_amyloid.csv",
                       na = c("", ".", "NA"))

```

 

