---
title: "p8105_hw2_abj2136"
output: github_document
date: "2026-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installing libraries
```{r message = FALSE}
library(tidyverse)
library(readxl)
library(readr)
```


# Problem 2

## Mr Trash Wheel
```{r}
# Looking at the sheets in the file
excel_sheets("202610 Trash Wheel Collection Data.xlsx")

#importing and cleaning the the mr trash wheel dataset

mr_trash_df = read_excel("202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Mr. Trash Wheel",
                      range = "A2:N742",
                      na = c("","NA", ".")
                      ) |>
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(sports_balls = as.integer(round(sports_balls))) |> 
  mutate(year = as.integer(year))
  
  # used glimpse in the console to determine variable types for all three data frames to 
  # determine which variables were entered as characters vs numeric values. In Mr.Trash 
  # the year variable needed to be converted to an integer, and in Professor trash the 
  # plastic bottles variable needed to be converted to numeric. 
```

## Professor Trash Wheel
```{r warning = FALSE}
#importing and cleaning the the prof trash wheel dataset

prof_trash_df = read_excel("202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Professor Trash Wheel",
                      range = "A2:M139",
                      na = c("","NA", ".")
                      ) |>
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(plastic_bottles = as.numeric(plastic_bottles))
```

## Gwynnda the Good Wheel of the West
```{r}
#importing and cleaning the the prof trash wheel dataset

gwynnda_df = read_excel("202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Gwynnda the Good Wheel of the W",
                      range = "A2:L394",
                      na = c("","NA", ".")
                      ) |>
  janitor::clean_names() |> 
  drop_na(dumpster)
```

## Merging datasets
```{r message = FALSE}
# creating a trash_wheel variable to keep track of which trash wheel is which after merging

mr_trash_df = mr_trash_df |> 
  mutate(trash_wheel = "Mr. Trash Wheel")

prof_trash_df = prof_trash_df |> 
  mutate(trash_wheel = "Professor Trash Wheel")

gwynnda_df = gwynnda_df |> 
  mutate(trash_wheel = "Gwynnda the Good Wheel of the W")

# merging the datasets

combined_trash_df = 
  bind_rows(
    mr_trash_df,
    prof_trash_df,
    gwynnda_df, 
  )

```

The final combined dataset contains `r nrow(combined_trash_df)` observations and `r ncol(combined_trash_df)` variables, such as `r names(combined_trash_df)`. Professor Trash Wheel collected a total of `r combined_trash_df |> filter(trash_wheel == "Professor Trash Wheel") |>  pull(weight_tons) |> sum(na.rm = TRUE)` tons of trash. Gwynnda the Good Wheel of the West collected `r combined_trash_df |> filter(trash_wheel == "Gwynnda the Good Wheel of the W", str_to_lower(month) == "june", year == 2022) |> pull(cigarette_butts) |> sum(na.rm = TRUE) |> scales::comma()` cigarette butts in June 2022. 


# Problem 3

## Importing and cleaning the baseline demographics data

```{r message = FALSE}

# keeping this here to see what original variables are called before skip

baseline_raw = read_csv("data_mci/MCI_baseline.csv")

# cleaning data set

baseline_df = read_csv("data_mci/MCI_baseline.csv",
                       na = c("", ".", "NA"),
                       skip = 1) |> 
  janitor::clean_names() |> 
  mutate(
    sex = recode_values(sex, 1 ~ "male", 0 ~ "female"),
    apoe4 = recode_values(apoe4, 1 ~ "carrier", 0 ~ "non-carrier")
    ) |> 
    filter(is.na(age_at_onset) | age_at_onset > current_age)
```

The first row of the data set acted like a code book, defining what the values meant for the variables below. I kept one read.csv line with the raw version so I could refer back to this code, as I skipped that row when I was cleaning my data. I had to recode the sex and apoe4 variables to change them from numeric to male/female and carrier/noncarrier. To filter the data to only keep those who had MCI at baseline, I kept everyone who had age of onset = na (never developed MCI) and everyone who developed MCI after baseline (based on if their age of MCI onset was greater than their age at baseline(current_age variable)), as these are the people who would be MCI free at baseline. 

`r nrow(baseline_raw)-1` participants were recruited into the study and `r nrow(baseline_df)` participants were eligible for the study. Of these participants `r baseline_df |> filter(!is.na(age_at_onset)) |> nrow()` developed MCI.The average baseline age among those who participated in the study was `r baseline_df |> pull(current_age) |> mean() |> round (1)` years. `r baseline_df |> filter(sex == "female") |> summarise(prop = mean(apoe4 == "carrier", na.rm = TRUE))|> pull(prop) |> scales::percent(accuracy = 0.1)` of the women in the study are APOE4 carriers.


## Importing and cleaning amyloid data
```{r message = FALSE}

 amyloid_df = read_csv("data_mci/mci_amyloid.csv",
                       na = c("", ".", "NA", "Na"),
                       skip = 1) |> 
  janitor::clean_names() |> 
  rename(id = study_id) |> 
  pivot_longer(
    baseline:time_8,
    names_to = "time_point",
    values_to = "amyloid_ratio"
  )
  
```
First, I cleaned the data by specifying missing values, skipping the first row 
with variable descriptions (so the first row was variables names), and cleaning
the names. Then, to make it possible to merge the amyloid and baseline data, I 
renamed the study_id variable in the amyloid data to match the id variable in 
the baseline data set.

This data set is wide, with multiple amyloid ratio measurements per person, 
at different time points. To tidy it and facilitate data analysis, I made 
it a longer data set, using pivot_longer to put all the amyloid_ratio values 
in one column (with multiple values for each patient id), with another variable 
called time_point to specify which timepoint the amyloid_ratio value came from.

The final tidied dataset contains `r nrow(amyloid_df)` observations. 

## Checking which participants are in only one dataset 

```{r message=FALSE}

# checking which participants appear in only one dataset

baseline_unique = anti_join(baseline_df, amyloid_df, by = "id")
amyloid_unique = anti_join(amyloid_df, baseline_df, by = "id")

# need to specify only the unique ids for the amyloid_df because each
# id has multiple rows
```

There are `r nrow(baseline_unique)` participants who appear only in the baseline 
data and not in the amyloid data. These may be people who dropped out of the 
study prior to biomarker collection. There are `r nrow(distinct(amyloid_unique, id))`
participants who appear only in the amyloid data set and not the baseline data.
These may be people who were excluded from the baselined dataset for having MCI
at baseline. 

## Merging the data sets 

```{r}

combined_mci_df =
  inner_join(baseline_df, amyloid_df, by = "id")

write_csv(combined_mci_df, "data_mci/combined_mci.csv")

```

In the final combined data set, there are `r nrow(combined_mci_df)` observations 
and the following variables: `r names(combined_mci_df)`. This data set only 
contains participants who appeared in both the baseline and amyloid data sets. 












